A simplified dynamic bioenergetic model for coral-algal symbiosis and coral bleaching

### 2016-10-19 JTB_ms/JTB_ms.tex
text updates

To analyze qualitative behavior, we ran the model to steady state across
gradients of external irradiance and nutrients (Fig. 2), which revealed
patterns consistent with observed phenomena in corals. Predicted growth
rates are low at low light and DIN (\textasciitilde{}0.01
d\textsuperscript{-1}), and begin increasing as both of these factors
increase (Fig. 2A). Low light is limiting to photosynthesis (Fig. 2D),
which leads to carbon-limitation of host growth (Fig. 2B) and an
associated increase in the symbiont to host biomass ratio (Fig. 2C). In
agreement with this trend are many observations of negative correlation
between irradiance and symbiont density (Stimson 1997; Brown et al.
1999; Fitt et al. 2000; Titlyanov et al. 2001). As higher light
alleviates light-limitation of photosynthesis, growth becomes less
carbon-limited. Similarly, elevating DIN up to \textasciitilde{}2 µM
alleviates nitrogen-limitation (Fig. 2B). Increased growth at higher DIN
is predicted by the DEB model of Muller et al. (2009), and has also been
observed experimentally (Muller-Parker et al. 1994; Tanaka et al. 2007;
Tanaka et al. 2013). However, DIN elevation beyond a certain point
(e.g., \textasciitilde{}3-4 µM in these simulations) has little effect
on growth as carbon becomes limiting (Fig. 2B). Very high nutrient
levels may even reduce growth (Shantz, Lemoine, and Burkepile 2015),
although these impacts are not likely to occur within the range of
concentrations considered here (\textless{}4 µM) (Ferrier-Pagès et al.
2000).

In addition to increasing growth, DIN also increases the symbiont to
host biomass ratio, a phenomenon also observed in reef corals (Marubini
and Davies 1996). S:H ratios also increase under low light (Stimson
1997; Fitt et al. 2000; Anthony and Hoegh-Guldberg 2003). At low DIN and
intermediate light, more typical of coral reef environments, symbiont to
host biomass ratios are around \textasciitilde{}0.1-0.3, which is
consistent with values reported in the literature (Muscatine, R
McCloskey, and E Marian 1981; Edmunds et al. 2011).

@@

The maximum predicted growth rates of \textasciitilde{}0.1
d\textsuperscript{-1}, occurring between \textasciitilde{}10-20 mol
photons m\textsuperscript{-2} s\textsuperscript{-1} light and
\textasciitilde{}4 µM DIN (Fig. 2A), are comparable to the rate of 0.07
d\textsuperscript{-1} measured by Tanaka et al. (2007) under similar
N-enriched conditions. Growth under conditions more typical of reef
environments (\textless{}0.5 µM DIN) are within the range of
\textasciitilde{}0.01-0.03 d\textsuperscript{-1}; these values are
slightly higher than most specific growth rates reported in the
literature, which are typically based on skeletal mass, and fall near or
below 0.01 d\textsuperscript{-1} (Osinga et al. 2011; Osinga et al.
2012) (though values as high as 0.025 d\textsuperscript{-1} have been
reported (Schutter et al. 2010)). These generally low skeletal growth
rates relative to predicted biomass growth may be explained by
differential substrate-limitation (i.e., CO\textsubscript{2} may limit
calcification, which is not present in the model); indeed, tissue and
skeletal growth may not always be well correlated (Anthony 2002).
Furthermore, bioerosion may reduce net skeletal growth, especially in
the field (Bak 1976). Nevertheless, when rates of increase in coral
tissue biomass have been directly measured over short periods, values
are consistent with model predictions (e.g., 0.07 d\textsuperscript{-1}
in Tanaka et al. (2007)), and similar rates (0.04 d\textsuperscript{-1})
have been measured in \emph{Aiptasia diaphana}, a non-calcifying
symbiotic anemone (Armoza-Zvuloni et al. 2014).

As irradiance continues to increase above \textasciitilde{}20 mol
photons m\textsuperscript{-2} d\textsuperscript{-1}, growth rates
decline until positive growth ceases above \textasciitilde{}35 µmol
photons m\textsuperscript{-2} d\textsuperscript{-1}. The mechanism
underlying this decline is the increase in light energy beyond the
capacity of photosynthesis and non-photochemical quenching: excess
excitation energy generates reactive oxygen species (ROS) (Weis 2008;
Roth 2014), which, in this model, have the phenomenological consequences
of reducing the photosynthetic rate (representing photoinhibition) and
increasing symbiont biomass loss (representing photodamage and/or
symbiont expulsion) (see Eynaud, Nisbet, and Muller 2011). Together,
these impacts reduce the symbiont to host biomass ratio (Fig. 2C), as
occurs during coral bleaching. This reduction in symbionts consequently
reduces the flux of fixed carbon to the host, resulting in increasing
carbon-limitation (Fig. 2B) and eventual cessation of growth (Fig. 2A).
Importantly, these consequences arise in the present model only when
light energy exceeds the capacity of photochemistry and NPQ, which in
turn depends on CO\textsubscript{2} availability. While previous models
framed photophysiological stress as a fixed response to absolute
irradiance levels (Eynaud, Nisbet, and Muller 2011), the current
implementation considers the dynamic balance of multiple energy sinks in
the causation of stress, which is more consistent with current
understanding of symbiosis dysfunction (Wooldridge 2013), and
establishes an important role of the host in providing
CO\textsubscript{2} for symbiont photosynthesis (Tansik, Fitt, and
Hopkinson 2015; Hopkinson, Tansik, and Fitt 2015).

While bleaching in response to high light alone has been observed
experimentally (Schutter et al. 2011; Downs et al. 2013), mass coral
bleaching events occur in response to high temperature (Hoegh-Guldberg
1999); thus, it is important to justify our consideration of light as
the primary stressor in the model. In reality, light and temperature
interact synergistically (Coles and Jokiel 1978; Jones et al. 1998), and
in fact, any stressor that disrupts the quenching of light energy may
lead to bleaching (Wooldridge 2010; Baker and Cunning 2015). This is
because the proximate cause of coral bleaching is excess excitation
energy, but the upstream events that lead to this situation may be
diverse: indeed, elevated temperature inhibits Rubisco functioning
(Jones et al. 1998) and repair of the D1 protein in photosystem II
(Warner, Fitt, and Schmidt 1999), which reduces the capacity of
photochemical quenching and leads to an excess of light energy. In this
way, elevated temperature serves to reduce the threshold above which
light causes stress to the system (Hoegh-Guldberg 1999); importantly,
light is still the proximate stressor. Therefore, we omitted temperature
from the model, which serves to maintain a desired level of simplicity,
while still allowing photophysiological stress and bleaching to be
simulated with biological realism in response to light. Thus, the
specific light levels that cause stress here should be interpreted in
this context, and the effects of high light are analagous to the effects
of high temperature.

The incorporation of light stress in the model sets an upper limit to
the amount of light at which a stable symbiotic interaction can be
maintained, but even below this threshold of breakdown, negative effects
of high light reduce steady state growth and symbiont:host biomass (Fig.
2A,C). This gradual decline is consistent with experimental results
showing that high light levels decrease growth (Schutter et al. 2011),
and field studies documenting optimum growth rates at intermediate
depths (Baker and Weber 1975; Huston 1985). Consequently, the model
predicts greater variation in state variables across light gradients
than was predicted by the model of Muller et al. (2009), which did not
include photoinhibition and photodamage. It is important to recognize
that these are steady state values predicted under a constant
environment at the given light and nutrient levels; a dynamic system may
temporarily cross this threshold and experience a period of symbiont
loss (bleaching) and reduced growth, after which a return to benign
conditions may restore symbiont biomass and positive growth. To explore
this further and illustrate the behavior of the model in more detail, we
evaluate a number of dynamic simulations below.

\subsection{Sensitivity analysis -- should come after the steady states
bc describes changes in steady
states}\label{sensitivity-analysis-should-come-after-the-steady-states-bc-describes-changes-in-steady-states}

Each default parameter value used in the model (Table 2) is based on
carefully selected relevant literature. A supplementary document
detailing the derivations of each of these default parameter values is
provided. Nevertheless, it is important to evaluate the sensitivity of
the model to changes in these parameter values. We measured fractional
change in steady state values in response to fractional change in
parameter values, relative to their default values (Table 2).

@@

@@

Therefore, we are highly interested in the conditions under which the
system switches from one state to another, representing the conditions
under which corals ``bleach'' due to progressive, reinforcing
carbon-limitation. Interestingly, this state can be reached in multiple
ways, by the acceleration of photoinhibition and photodamage, or by high
external DIN, which causes the symbiont to share less carbon,
subsequently causing its own CO2-limitation, and progressive
C-limitation of the system. \#\# varying light: consequences of excess
light and light-limitation \#\# varying DIN: role of N-limitation and
consequences of elevated N \#\# elevated nutrients cause system to
become C-limited (photo: CO2 limited; host: fixed-C limited) \#\#
C-limitation increases susceptibility to bleaching \#\# consequences of
a less efficient but more tolerant symbiont \#\# Rate of symbiont
repopulation (Berner et al. 1993) \#\# Bicarbonate addition promotes
coral growth (Marubini and Thake 1999) \#\# Seasonal change in sym.
density is larger when nutrients are elevated (Stimson 1997). Such a
phenomenon was predicted by Fitt et al. 2000. \#\# Death rule: negative
growth for a certain amount of time should be defensible.

@@

\hypertarget{ref-Anthony:2003p3632}{}
Anthony, Kenneth R N, and Ove Hoegh-Guldberg. 2003. ``Variation in coral
photosynthesis, respiration and growth characteristics in contrasting
light microhabitats: an analogue to plants in forest gaps and
understoreys?'' \emph{Functional Ecology} 17 (April): 246--59.

@@

>>>>

@@

To analyze qualitative behavior, we ran the model to steady state across
gradients of external irradiance and nutrients (Fig. 2), which revealed
patterns consistent with observed phenomena in corals. Predicted growth
rates are low at low light and DIN (\textasciitilde{}0.01
d\textsuperscript{-1}), and begin increasing as both of these factors
increase (Fig. 2A). Low light limits photosynthetic rates, resulting in
less fixed carbon shared with the host and an associated increase in the
symbiont to host biomass ratio (Fig. 2B). In agreement with this trend
are many observations of negative correlation between irradiance and
symbiont density (Stimson 1997; Brown et al. 1999; Fitt et al. 2000;
Titlyanov et al. 2001). As higher light alleviates light-limitation of
photosynthesis, growth becomes less carbon-limited.

Similarly, increasing DIN up to \textasciitilde{}2 µM alleviates
nitrogen-limitation (Fig. 2A). Increased growth at higher DIN is
predicted by the DEB model of Muller et al. (2009), and has also been
observed experimentally (Muller-Parker et al. 1994; Tanaka et al. 2007;
Tanaka et al. 2013). However, DIN elevation beyond a certain point
(e.g., \textasciitilde{}3-4 µM in these simulations) has little effect
on growth as carbon becomes limiting. Very high nutrient levels may even
reduce growth (Shantz, Lemoine, and Burkepile 2015), although these
impacts are not likely to occur within the range of concentrations
considered here (\textless{}4 µM) (Ferrier-Pagès et al. 2000). In
addition to increasing growth, DIN also increases the symbiont to host
biomass ratio (Fig. 2B), a phenomenon also observed in reef corals
(Marubini and Davies 1996). At low DIN and intermediate light, more
typical of coral reef environments, symbiont to host biomass ratios are
around \textasciitilde{}0.1-0.3, which is consistent with values
reported in the literature (Muscatine, R McCloskey, and E Marian 1981;
Edmunds et al. 2011; Hawkins et al. 2016).

The maximum predicted growth rates of \textasciitilde{}0.1
d\textsuperscript{-1}, occurring between \textasciitilde{}10-20 mol
photons m\textsuperscript{-2} s\textsuperscript{-1} light and
\textasciitilde{}4 µM DIN (Fig. 2A), are comparable to the rate of 0.07
d\textsuperscript{-1} measured by Tanaka et al. (2007) under similar
N-enriched conditions. Predicted growth rates under conditions more
typical of reef environments (\textless{}0.5 µM DIN) are
\textasciitilde{}0.01-0.03 d\textsuperscript{-1}. Observed specific
growth rates, which are typically based on skeletal mass, fall near or
below 0.01 d\textsuperscript{-1} (Osinga et al. 2011; Osinga et al.
2012) (though values as high as 0.025 d\textsuperscript{-1} have been
reported (Schutter et al. 2010)). That predicted growth rates are
slightly higher than most observations is not surprising, as the model
does not account for external ecological factors that may reduce growth,
such as competition, predation, and bioerosion. Furthermore, the model
does not explicitly account for skeletal formation, which may not always
be well correlated with biomass growth, potentially due to differential
substrate-limitation (Anthony 2002). Nevertheless, when rates of
increase in coral tissue biomass have been directly measured over short
periods, values are consistent with model predictions (e.g., 0.07
d\textsuperscript{-1} in Tanaka et al. (2007)), and similar rates (0.04
d\textsuperscript{-1}) have been measured in \emph{Aiptasia diaphana}, a
non-calcifying symbiotic anemone (Armoza-Zvuloni et al. 2014).

As irradiance continues to increase above \textasciitilde{}20 mol
photons m\textsuperscript{-2} d\textsuperscript{-1}, growth rates
decline until positive growth ceases above \textasciitilde{}35 µmol
photons m\textsuperscript{-2} d\textsuperscript{-1} (Fig. 2A). The
mechanism underlying this decline is the increase in light energy beyond
the capacity of photosynthesis and non-photochemical quenching: excess
excitation energy generates reactive oxygen species (ROS) (Weis 2008;
Roth 2014), which, in this model, have the phenomenological consequences
of reducing the photosynthetic rate (representing photoinhibition) and
increasing symbiont biomass loss (representing photodamage and/or
symbiont expulsion) (see Eynaud, Nisbet, and Muller 2011). Together,
these impacts reduce the symbiont to host biomass ratio (Fig. 2B), as
occurs during coral bleaching. This reduction in symbionts consequently
reduces the flux of fixed carbon to the host, resulting in increasing
carbon-limitation (Fig. 2B) and eventual cessation of growth (Fig. 2A).

While bleaching in response to high light alone has been observed
experimentally (Schutter et al. 2011; Downs et al. 2013), mass coral
bleaching events occur in response to high temperature (Hoegh-Guldberg
1999); thus, it is important to justify our consideration of light as
the primary stressor in the model. In reality, light and temperature
interact synergistically (Coles and Jokiel 1978; Jones et al. 1998), and
in fact, any stressor that disrupts the quenching of light energy may
lead to bleaching (Wooldridge 2010; Baker and Cunning 2015). This is
because the proximate cause of coral bleaching is excess excitation
energy, but the upstream events that lead to this situation may be
diverse: indeed, elevated temperature may inhibit Rubisco functioning
(Jones et al. 1998) and the repair of the D1 protein in photosystem II
(Warner, Fitt, and Schmidt 1999), which reduces the capacity of
photochemical quenching and leads to an excess of light energy. In this
way, elevated temperature serves to reduce the threshold above which
light causes stress to the system (Hoegh-Guldberg 1999); importantly,
light is still the proximate stressor. Therefore, we omitted temperature
from the model to maintain a desired level of simplicity, while still
allowing photophysiological stress and bleaching to be simulated with
biological realism in response to light. Thus, while the model is
currently accurate from a conceptual and mechanistic standpoint, its
value as a predictive tool under scenarios of changing temperature will
require further development.

The incorporation of light stress in the model sets an upper limit to
the amount of light at which a stable symbiotic interaction can be
maintained, but even below this threshold of breakdown, negative effects
of high light reduce steady state growth and symbiont:host biomass (Fig.
2). This gradual decline is consistent with experimental results showing
that high light levels decrease growth (Schutter et al. 2011), and field
studies documenting optimum growth rates at intermediate depths (Baker
and Weber 1975; Huston 1985). By incorporating these impacts of light
stress, the model predicts greater, and more realistic, variation in
state variables across light gradients than was predicted by the model
of Muller et al. (2009), which did not include photoinhibition and
photodamage.

It is important to recognize that the boundary set by photo-oxidative
stress on a stable symbiosis under steady state conditions (Fig. 2) may
be temporarily crossed by a dynamic system, which may experience a
period of symbiont loss (bleaching) and reduced growth, after which a
return to benign conditions may restore symbiont biomass and positive
growth. To explore this further and illustrate the behavior of the model
in more detail, we evaluate a number of dynamic simulations below (see
``Dynamic behavior'').

\subsection{Sensitivity analysis}\label{sensitivity-analysis}

Each default parameter value used in the model (Table 2) is based on
carefully selected relevant literature. The derivations of each of these
default parameter values are described in the Supplementary Information.
Nevertheless, it is important to evaluate the sensitivity of the model
to changes in these parameter values. Moreover, these sensitivity plots
serve to demonstrate the behavior of the dynamical system. We measured
fractional change in steady state values in response to fractional
changes in parameter values, relative to their default values, under
environmental conditions typical of coral reefs.

@@

Overall, relative changes in the steady state of the system were less
than the equivalent relative change in parameter value, indicating that
the system maintains internal stability. However, changes in certain
parameter values have greater impacts: increasing \(j_{Nm}\) or
decreasing \(K_N\) both dramatically increase host growth, demonstrating
the strong nitrogen-limitation that characterizes these symbioses (REF).
The parameterr \(\bar{a}^*\) has a strong impact on S:H biomass ratios,
because this parameter determines the amount of light absorbed by
symbionts, thus low values of this parameter increase light-limitation
of the system. Altering the maximum growth and turnover rates of
symbiont and host biomass also have expected impacts on steady states.
Parameters having to do with photooxidative stress and bleaching have
little impact under these environmental conditions, but of course have
larger impacts under higher light (e.g., Fig. S6).

@@

Therefore, we are highly interested in the conditions under which the
system switches from one state to another, representing the conditions
under which corals ``bleach'' due to progressive, reinforcing
carbon-limitation. Interestingly, this state can be reached in multiple
ways, by the acceleration of photoinhibition and photodamage, or by high
external DIN, which causes the symbiont to share less carbon,
subsequently causing its own CO2-limitation, and progressive
C-limitation of the system. \#\# varying light: consequences of excess
light and light-limitation \#\# varying DIN: role of N-limitation and
consequences of elevated N \#\# elevated nutrients cause system to
become C-limited (photo: CO2 limited; host: fixed-C limited) \#\#
C-limitation increases susceptibility to bleaching Importantly, these
negative impacts of high irradiance only arise when light energy exceeds
the capacity of photochemistry and NPQ, which in turn depends on
CO\textsubscript{2} availability. While previous models framed
photophysiological stress as a fixed response to absolute irradiance
levels (Eynaud, Nisbet, and Muller 2011), the current implementation
considers the dynamic balance of multiple energy sinks in the causation
of stress, which is more consistent with current understanding of
symbiosis dysfunction (Wooldridge 2013), and establishes an important
role of the host in providing CO\textsubscript{2} for symbiont
photosynthesis (Tansik, Fitt, and Hopkinson 2015; Hopkinson, Tansik, and
Fitt 2015). \#\# consequences of a less efficient but more tolerant
symbiont \#\# Rate of symbiont repopulation (Berner et al. 1993) \#\#
Bicarbonate addition promotes coral growth (Marubini and Thake 1999)
\#\# Seasonal change in sym. density is larger when nutrients are
elevated (Stimson 1997). Such a phenomenon was predicted by Fitt et al.
2000. \#\# Death rule: negative growth for a certain amount of time
should be defensible.

@@

@@

\hypertarget{ref-Hawkins:2016jj}{}
Hawkins, Thomas D, Julia C G Hagemeyer, Kenneth D Hoadley, Adam G Marsh,
and Mark E Warner. 2016. ``Partitioning of Respiration in an
Animal-Algal Symbiosis: Implications for Different Aerobic Capacity
between Symbiodinium spp.'' \emph{Frontiers in Physiology} 7 (April).
Frontiers: 125.
doi:\href{https://doi.org/10.3389/fphys.2016.00128}{10.3389/fphys.2016.00128}.
